Single Bivalent Level Analysis

### 2020-01-23 doc/Bivalent_Analysis.Rmd
added a fig for sis-co-ten example

#check for dup
length(duplicated(BivData_from_15biv_cells$Obj.ID))#2605
anyDuplicated(BivData_from_15biv_cells$Obj.ID)#1997
dups.dummy <- BivData_from_15biv_cells[duplicated(BivData_from_15biv_cells$Obj.ID),]

@@

real.long.bivData <- BivData_from_15biv_cells[BivData_from_15biv_cells$long.Biv == "yes",]
table(real.long.bivData$fileName)

@@

Well fixing those bigs took forever

@@

@@

@@

@@

I used the mixed model with the lme()

@@

```{r MM_2_SC.length, echo=FALSE}

@@

#can't make the 
SC_length.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

anova(SC_length.MM)

thing1  <- summary(SC_length.MM)
thing2 <- drop1(SC_length.MM, test="Chisq")

rand_SC.Length_MM <- exactRLRT(SC_length.MM)
rand_SC.Length_MM

#this pvalue for ran is 0.0027 -- the strain effect 
```

@@

@@

#the summary for both these models is the same ... should I be worried?

@@

@@

@@

@@

#remove the SKIVE male line
mouse.avs_4MM <- mouse.avs_4MM[!(is.na(mouse.avs_4MM$mean_IFD.2CO_ABS) | mouse.avs_4MM$mean_IFD.2CO_ABS==""), ]

@@

#make

@@

summary(log.reg_IFD.ASB)

log.reg_IFD.PER <- glm(Rec.group ~ mean_IFD.2CO_PER, 
                            data=mouse.avs_4MM_M.male, family=binomial(link="logit"))
summary(log.reg_IFD.PER)

@@

@@

#clean up all the non 1 to 4 hand.foci counts
MELT_Curated_BivData <- MELT_Curated_BivData[MELT_Curated_BivData$hand.foci.count <5, ]
# remove the na's
MELT_Curated_BivData <- MELT_Curated_BivData[!(is.na(MELT_Curated_BivData$hand.foci.count) | MELT_Curated_BivData$hand.foci.count==""), ]

Musc.female_MELT_curBivData <- MELT_Curated_BivData[ ((MELT_Curated_BivData$sex == "female") & (MELT_Curated_BivData$subsp == "Musc") ), ]

CO.Pos.density_Musc_female <- ggplot(Musc.female_MELT_curBivData, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Musc female \nNormalized Foci Positions")
CO.Pos.density_Musc_female

#Dom female
Dom.female_MELT_curBivData <- MELT_Curated_BivData[ ((MELT_Curated_BivData$sex == "female") & (MELT_Curated_BivData$subsp == "Dom") ), ]

CO.Pos.density_Dom_female <- ggplot(Dom.female_MELT_curBivData, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Dom female \nNormalized Foci Positions")
CO.Pos.density_Dom_female

#Musc male
Musc.male_MELT_curBivData <- MELT_Curated_BivData[ ((MELT_Curated_BivData$sex == "male") & (MELT_Curated_BivData$subsp == "Musc") ), ]

CO.Pos.density_Musc_male <- ggplot(Musc.male_MELT_curBivData, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Musc male \nNormalized Foci Positions")
CO.Pos.density_Musc_male

#Dom male
Dom.male_MELT_curBivData <- MELT_Curated_BivData[ ((MELT_Curated_BivData$sex == "male") & (MELT_Curated_BivData$subsp == "Dom") ), ]

CO.Pos.density_Dom_male <- ggplot(Dom.male_MELT_curBivData, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Dom male \nNormalized Foci Positions")
CO.Pos.density_Dom_male

@@

Split up the plot data by subsp - sex. 
Try adding in the numbers of observations?
think about adding a vertical line for centromere, for the position means. Think about removing the extra Musc strains.

@@

@@

@@

```{r CO.pos.poly}
#I
#CO_pos

@@

need more MSM male curated measures!

>>>>

#check for dup
#length(duplicated(BivData_from_15biv_cells$Obj.ID))#2605
#anyDuplicated(BivData_from_15biv_cells$Obj.ID)#1997
dups.dummy <- BivData_from_15biv_cells[duplicated(BivData_from_15biv_cells$Obj.ID),]

@@

real.long.bivData <- BivData_from_15biv_cells[BivData_from_15biv_cells$long.Biv == "yes",]
#table(real.long.bivData$fileName)

@@

Well fixing those bigs took forever. Use the DF, **real.long.bivData** for comparing the longest bivalents.  still v unblancd across sexes and strains ~ merge with the manual whole cell biv measures later...

MSM male only has data from 1 cell, and PWD 3, uh oh -- may need to merge sooner.

First proof of principl will be comparisons within category (long vs all)

```{r mouse.av.long.bivData, echo=FALSE}
#real.long.bivData

#calq the mouse averages for

Long_biv_mouse.avs_4MM <- ddply(real.long.bivData, c("mouse"), summarise,
                        Nmice = length(unique(mouse)),
                        Ncells  = length(Obj.ID),
                        mean_SC = as.numeric(format(round(  mean(chromosomeLength), 3 ), nsmall=3) ),
                        cV = cv(chromosomeLength),
                        var = as.numeric(format(round(   var(chromosomeLength),3), nsmall=3) ),
                        sd   = round(sd(chromosomeLength), 3),
                        se   = round(sd / sqrt(Ncells), 3),
                       #IFD things
                #  mean_IFD_ABS = as.numeric(format(round(  mean(IFD1_ABS, na.rm = TRUE), 3 ), nsmall=3) ),
                # mean_IFD_PER = as.numeric(format(round(  mean(IFD1_PER, na.rm = TRUE), 3 ), nsmall=3) ),

Long_biv_mouse.avs_4MM <- add_sex(Long_biv_mouse.avs_4MM)
Long_biv_mouse.avs_4MM <- add_strain(Long_biv_mouse.avs_4MM)
Long_biv_mouse.avs_4MM <- add_subsp(Long_biv_mouse.avs_4MM)

in code chunk above I ran the mouse averages for the longest bivalets (680 bivalents, from 54 mice. full dataset is 10202 bivalents from 86 mice.

@@

# Adjusting for XX

The female mouse averages should have adjustments for the XX. working on code to estimate the SC length from 3rd largest bivalent from female whole cell data across strains. Subtract this amount from the female mouse averages ... This isn't the best solution -- since I can't determine what proportion of cells for female mouse averages include the XX, (most cells are missing at least 3 bivalents)

- Of all female single bivalents 5% are XXs (1 of 20).

- The XX is large, thought to be within the top 25% longest bivalents of the cell (3rd largest by Mb).

- What is the average % of XX for whole cell SC (sum(all bivalents)). Lets guess 12% of a cell's total SC area is XX.

- do the algebra...

@@

remeber the long biv data set in **real.long.bivData**

@@

#model for long Bivs
long_bivData_SC.Length_MM <- lme(mean_SC ~ subsp * sex, data= Long_biv_mouse.avs_4MM, random=list(strain=pdDiag(~sex) ) )
as.data.frame( round(  summary(long_bivData_SC.Length_MM)$tTable, 6 ) )

@@

I used the mixed model with the lme() for the full Curated BivData and long bivdata

@@

```{r MM_2_SC.length, echo=FALSE}
#Long_biv_mouse.avs_4MM

@@

#can't make the 
SC_length.lmer.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

anova(SC_length.lmer.MM)

thing1  <- summary(SC_length.lmer.MM)
thing2 <- drop1(SC_length.lmer.MM, test="Chisq")

rand_SC.Length_MM <- exactRLRT(SC_length.lmer.MM)
rand_SC.Length_MM

#this pvalue for ran is 0.0027 -- the strain effect

##long bivalents
long.bivs_SC_length.lmer.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=Long_biv_mouse.avs_4MM, REML = TRUE)

anova(long.bivs_SC_length.lmer.MM) 
#sex is larges explantion of variance

thing1  <- summary(long.bivs_SC_length.lmer.MM)
thing2 <- drop1(long.bivs_SC_length.lmer.MM, test="Chisq")

rand_SC.Length_MM <- exactRLRT(long.bivs_SC_length.lmer.MM)
rand_SC.Length_MM

#strain no sig, but underpowered

@@

- The Long Biv Data set largely agrees with the full curated dataset

- I caveat I haven't addressed yet, is the XX in the female Biv data averages ---

@@

summary(log.reg_SC.length_Musc)

@@

```{r long.sc_log.res_meanSC}

long.SC.mouse.avs_4MM_male <- Long_biv_mouse.avs_4MM[Long_biv_mouse.avs_4MM$sex == "male",]
long.SC.mouse.avs_4MM_Mmale <- long.SC.mouse.avs_4MM_male[long.SC.mouse.avs_4MM_male$subsp == "Musc",]

#divide musc males into groups
long.SC.mouse.avs_4MM_male$Rec.group <- ifelse(grepl("PWD", long.SC.mouse.avs_4MM_male$strain), 1, 
                                    ifelse(grepl("MSM", long.SC.mouse.avs_4MM_male$strain), 1, 0))

long.biv.log.reg <- glm(Rec.group ~ mean_SC, 
                            data=long.SC.mouse.avs_4MM_male, family=binomial(link="logit"))

summary(long.biv.log.reg)#NS effect (likelt underpowered)
```

@@

ToDo fix the function to calculat IFD2, IFD3

@@

```{r blah.again}
#long SC Long_biv_mouse.avs_4MM
#real.long.bivData

#Only use 2CO bivalents
real.long.bivData.22 <- real.long.bivData[real.long.bivData$hand.foci.count == 2,]
real.long.bivData.22 <- real.long.bivData.22[(!is.na(real.long.bivData.22$hand.foci.count)),]

long.biv_IFD_look_ABS <- ggplot(real.long.bivData.22, aes(y = IFD1_ABS, x= mouse, color=strain))+geom_boxplot(alpha=.2)+
  geom_jitter(alpha=.2)+
  ggtitle("Interfocal Distances, ABS")+
  scale_color_manual(values=colors_of_strains)+  facet_wrap(sex~subsp, scales = "free")+ylim(c(20,160))+theme(legend.position="none")

long.biv_IFD_look_ABS

long.biv_IFD_look_PER <- ggplot(real.long.bivData.22, aes(y = IFD1_PER, x= mouse, color=strain))+geom_boxplot(alpha=.2)+
  geom_jitter(alpha=.2)+
  ggtitle("Interfocal Distances, ABS")+
  scale_color_manual(values=colors_of_strains)+  facet_wrap(sex~subsp, scales = "free")+ylim(c(0,1))+theme(legend.position="none")

long.biv_IFD_look_PER

@@

#remove the SKIVE male line which is na
mouse.avs_4MM <- mouse.avs_4MM[!(is.na(mouse.avs_4MM$mean_IFD.2CO_ABS) | mouse.avs_4MM$mean_IFD.2CO_ABS==""), ]

@@

#make t.tests for pooled high and low

@@

 #summary(log.reg_IFD.ASB)

log.reg_IFD.PER <- glm(Rec.group ~ mean_IFD.2CO_PER, 
                            data=mouse.avs_4MM_M.male, family=binomial(link="logit"))
 #summary(log.reg_IFD.PER)

@@

for the long chrm, pull the object ids out of the melt table.

### Normalized 1CO positions

@@

#clean up all the non 1 to 4 hand.foci counts
MELT_Curated_BivData.1CO <- MELT_Curated_BivData[MELT_Curated_BivData$hand.foci.count < 2, ]

MELT_Curated_BivData.1CO <- MELT_Curated_BivData.1CO[MELT_Curated_BivData.1CO$PER.Foci.Pos < 1.01, ]

# remove the na's
MELT_Curated_BivData.1CO <- MELT_Curated_BivData.1CO[!(is.na(MELT_Curated_BivData.1CO$hand.foci.count) | MELT_Curated_BivData.1CO$hand.foci.count==""), ]

CO.Pos.density_1CO <- ggplot(MELT_Curated_BivData.1CO, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(sex~subsp, ncol=2)+ggtitle("Normalized 1CO Foci Positions")
CO.Pos.density_1CO

#category specific plots, don't run for now.
Dom.female_MELT_curBivData <- MELT_Curated_BivData[ ((MELT_Curated_BivData$sex == "female") & (MELT_Curated_BivData$subsp == "Dom") ), ]
CO.Pos.density_Dom_female <- ggplot(Dom.female_MELT_curBivData, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Dom female \nNormalized Foci Positions")
#Musc male
Musc.male_MELT_curBivData <- MELT_Curated_BivData[ ((MELT_Curated_BivData$sex == "male") & (MELT_Curated_BivData$subsp == "Musc") ), ]

CO.Pos.density_Musc_male <- ggplot(Musc.male_MELT_curBivData, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Musc male \nNormalized Foci Positions")
#Dom male
Dom.male_MELT_curBivData <- MELT_Curated_BivData[ ((MELT_Curated_BivData$sex == "male") & (MELT_Curated_BivData$subsp == "Dom") ), ]

CO.Pos.density_Dom_male <- ggplot(Dom.male_MELT_curBivData, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Dom male \nNormalized Foci Positions")

@@

Above plot focuses on the 1CO bivalent normalized positions -- since CO interference controls the general position of COs when there are multiple COs. This plot shows the sexual dimorphism in the density plots.

consider adding annotate_text for the number of observations in each category.
think about adding a vertical line for centromere, for the position means. Think about removing the extra Musc strains.

```{r norm.CO.pos_longBiv, echo=FALSE, eval=FALSE}
#real.long.bivData
#MELT_Curated_BivData

MELT_long_biv <- MELT_Curated_BivData[ MELT_Curated_BivData$Obj.ID %in% real.long.bivData$Obj.ID, ]

MELT_long_biv <- add_sex(MELT_long_biv)
MELT_long_biv <- add_strain(MELT_long_biv)
MELT_long_biv <- add_subsp(MELT_long_biv)
MELT_long_biv <- add_category(MELT_long_biv)

long.biv_Dom.male <- MELT_long_biv[ (MELT_long_biv$subsp == "Dom") & (MELT_long_biv$sex == "male"), ]

Dom.male_MELT_long <- ggplot( long.biv_Dom.male, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Dom male long biv\nNormalized Foci Positions")
Dom.male_MELT_long

long.biv_Musc.male <- MELT_long_biv[ (MELT_long_biv$subsp == "Musc") & (MELT_long_biv$sex == "male"), ]

Musc_male_MELT_long <- ggplot( long.biv_Musc.male, aes(x=PER.Foci.Pos, fill=strain) ) + geom_density(alpha=.2) + facet_wrap(~hand.foci.count, ncol=1)+ggtitle("Musc male long biv\nNormalized Foci Positions")
Musc_male_MELT_long

#there are v few observations in this dataset

@@

Tired runnin plots for the long biv data set, I think there are too few observations.

add boxplots for

```{r CO1_foci_pos_boxplot, echo=FALSE}

CO.Pos.boxplots_1CO <- ggplot(MELT_Curated_BivData.1CO, aes(x= strain, y=PER.Foci.Pos, fill=strain) ) + geom_boxplot(alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("boxy plots")+geom_jitter(alpha=.1)
CO.Pos.boxplots_1CO

These bosplots show that females have a much more medial poition of single foci bivalents, (much closer to 50% compared to males). They also show that Musc males' Foci1 position is slightly more central / medial compared to the same type of positions in the Dom male strains. MOLF males have much more medial positions than other strains.

##### Mixed models norm 1CO positions

@@

### Siscoten

The logic of how the sis-co-ten metric is outlined in the figure below. The authors from this paper call it by something else. looking at this siscoten metric is a way to model different cohesin outcomes as a consequence of different numbers and placements of chiasmata/CO. This metric is calculated from SC area -- which we are using as a proxy to the amount of cohesion at metaphase.

![Logic of Sis-co-ten metric, from (Lee, J. (2019). Is age-related increase of chromosome segregation errors in mammalian oocytes caused by cohesin deterioration?. Reproductive Medicine and Biology.)](figureFiles/Lee2019_Fig2.png)

Males have much clearer seperation of siscoten across chrm classes. This is emphasized when SC length is also plotted.

### Telomere and centromere Distance

My metric for telomere and centromere distance measure the distance of the nearest foci to the ends of the bivalent (SC). In the plots below each point is a single bivalent. I choose not to use the mark for centromere because it seems noisy and inconsistant...

```{r telo.dist, echo=FALSE}

sis.co.ten.plot_WSB_female <- ggplot(data = Curated_BivData[Curated_BivData$category == "WSB female",], aes(y=SisCoTen, x=chromosomeLength, color = as.factor(hand.foci.count) ))+
  geom_point()+facet_wrap(sex~strain)

Curated_BivData <- Curated_BivData[Curated_BivData$hand.foci.count < 5,]
Curated_BivData <- Curated_BivData[!(is.na(Curated_BivData$hand.foci.count) | Curated_BivData$hand.foci.count==""), ]

telo.dist_boxplots <- ggplot(Curated_BivData, aes(x= hand.foci.count, y=telo_dist ) )+ geom_boxplot(aes( fill=hand.foci.count, group=strain), alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("boxy plots")+ ylim(c(0, 150))

telo_jitter <- ggplot(Curated_BivData, aes(x= strain, y=telo_dist, color=as.factor(hand.foci.count) ) )+ geom_jitter(alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("raw telomere distance")+ ylim(c(0, 150))
telo_jitter

nrm.telo_jitter <- ggplot(Curated_BivData, aes(x= strain, y=telo_dist_PER, color=as.factor(hand.foci.count) ) )+ geom_jitter(alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("Normalized telomere distance")+ylim(c(0,1.2))
nrm.telo_jitter

#why do a different number of rows removed each time??

Males on average have much lower raw telomere distance (reflects the telomere bias) compared to females. In Males, 2CO bivalents have very low telomere distances, while the 1CO bivalents have a greater range. In females the ranges of telomere distances have much more overlap.

```{r cent.dist, echo=FALSE}

cent.dist_boxplots <- ggplot(Curated_BivData, aes(x= strain, y=dis.cent ) )+ geom_boxplot(aes( fill=hand.foci.count, group=strain), alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("raw centromere distance")+ ylim(c(0, 150))

cent.dist_jitter <- ggplot(Curated_BivData, aes(x= strain, y=dis.cent, color=as.factor(hand.foci.count) ) )+ geom_jitter(alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("raw centromere dist")+ ylim(c(0, 200))
cent.dist_jitter

norm.cent.dist_jitter <- ggplot(Curated_BivData, aes(x= strain, y=dis.cent.PER, color=as.factor(hand.foci.count) ) )+ geom_jitter(alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("Normalized centromere distance")+ylim(c(0,1.2))
norm.cent.dist_jitter

The normalized centromere plots show that in Musc males, on 2CO bivalents the 1st CO is closer to the centromere end than in Dom males.

Females have more overlap in the distributions of centromere distances across chromosome class compared to males.

@@

```{r CO.pos.poly}

#CO_pos

@@
